---
title: "MIC analysis of Methicillin-resistant Staphylococcus epidermidis"
author: "Olivier Restif, with data from Jesper Larsen"
date: "21/04/2022"
output:
  html_document: default
  pdf_document: default
---

Last updated: `r Sys.Date()`

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
library(tidyverse)
library(readxl)
library(mixtools)
library(GGally)
source("../MIC_mixture_models_v3.R")
```

```{r}
mic.data <- read_excel("S epidermidis MIC data.xlsx",1) %>%
	mutate(AC.CA_2.1 = str_split(`AC/CA (2:1 in MHB)`,"/",simplify = T)[,2] %>% as.numeric(),
		 AC.CA_15 = ifelse(`AC/CA (15)`=="≦0.0039",0.0039,`AC/CA (15)`) %>% as.numeric(),
		 Clavulanic_acid = ifelse(`Clavulanic acid`==">256",Inf,`Clavulanic acid`) %>% as.numeric(),
		 .keep= "unused")

mic.data.long <- mic.data %>% pivot_longer(cols=3:8, names_to = "Antibiotic",values_to = "MIC")
```

# Data description

Number of samples: `r nrow(mic.data)`

```{r}
mic.data %>% group_by(Genotype_EMH) %>% summarise(Samples=n())
```


```{r, out.width="120%"}
ggplot(mic.data.long, aes(x=as.factor(MIC))) +
	geom_bar(aes(fill = Genotype_EMH)) +
	theme(axis.text.x = element_text(angle=75,hjust=1)) +
	facet_wrap(vars(Antibiotic))
```

Pairwise correlations. Values show log2(MIC).

```{r}
cor(mic.data[,3:8],method = "spearman") %>% round(3)
```



```{r}
ggpairs(mic.data.long %>% 
	  	mutate(log2.MIC = ifelse(is.finite(MIC),round(log2(MIC)),9)) %>%
	  	select(-MIC) %>%
	  	pivot_wider(names_from = Antibiotic, values_from = log2.MIC),
	  columns = 3:8,
	  diag=list(continuous="barDiag")
	  ) 
```


```{r}
ggpairs(mic.data.long %>% 
	  	mutate(log2.MIC = ifelse(is.finite(MIC),round(log2(MIC)),9)) %>%
	  	select(-MIC) %>%
	  	pivot_wider(names_from = Antibiotic, values_from = log2.MIC),
	  columns = 3:8,
	  ggplot2::aes(colour=Genotype_EMH),
	  diag=list(continuous="barDiag")
	  ) 
```

Spearman correlation coefficients:
```{r}
ggpairs(mic.data.long %>% 
	  	mutate(log2.MIC = ifelse(is.finite(MIC),round(log2(MIC)),9)) %>%
	  	select(-MIC) %>%
	  	pivot_wider(names_from = Antibiotic, values_from = log2.MIC),
	  columns = 3:8,
	  upper = list(continuous = wrap("cor", method = "spearman")),
	  diag=list(continuous="barDiag"),
	  title="Spearman correlation coefficients"
	  ) 

```


# Mixture models

## AC/CA (15)

```{r}
ggplot(mic.data.long %>% filter(Antibiotic=="AC.CA_15"), aes(x=as.factor(MIC))) +
	geom_bar(aes(fill = Genotype_EMH)) +
	theme(axis.text.x = element_text(angle=75,hjust=1))
```


```{r, include=F}
AC.CA_15.range <- read_excel("S epidermidis MIC data.xlsx",3)$`AC/CA (15)` %>% sort()
AC.CA_15.MM.mle <- fit.multi.mix(mic.data$AC.CA_15, AC.CA_15.range, K=3)
```

```{r}
AC.CA_15.MM.mle$cut.offs
```

Proposed ECOFF for AC/CA 15: 5.66.

- The 71 isolates with MIC $\leq 4$ are labelled "susceptible".
- The 67 isolates with MIC $\geq 8$ are labelled "resistant".


Predicted distribution of susceptible (S) and resistant (R) isolates from the model, with actual data shown in grey:

```{r}
ggplot(AC.CA_15.MM.mle$pred.n[[3]] %>% 
	 	mutate(MIC=factor(High), Pred.S = round(Pred.1 + Pred.2,1), Pred.R = round(Pred.3,1), .keep="none") %>%
	 	pivot_longer(2:3,names_to="Category", values_to="N")) + 
	geom_col(aes(x=MIC,y=N, fill=Category)) +
	geom_vline(xintercept=11.5)+
	theme(axis.text.x = element_text(angle=75,hjust=1)) +
	geom_col(data=AC.CA_15.MM.mle$pred.n[[3]] %>% transmute(MIC=factor(High), N=N), aes(x=MIC, y=N), fill=NA, col=grey(0.5))
	
```


## AC/CA (2:1 in MHB)

Here we only report the MIC for the second drug, e.g. 4 for "8/4".

```{r}
ggplot(mic.data.long %>% filter(Antibiotic=="AC.CA_2.1"), aes(x=as.factor(MIC))) +
	geom_bar(aes(fill = Genotype_EMH)) +
	theme(axis.text.x = element_text(angle=75,hjust=1))
```

```{r, include=F}
AC.CA_2.1.range <- read_excel("S epidermidis MIC data.xlsx",3)$`AC/CA (2:1 in MHB)` %>% sort()
AC.CA_2.1.MM.mle <- fit.multi.mix(mic.data$AC.CA_2.1, AC.CA_2.1.range, K=3)
```

```{r}
AC.CA_2.1.MM.mle$cut.offs
```

Proposed ECOFF for "AC/CA (2:1 in MHB)": 11.3/5.65.

- The 69 isolates with MIC $\leq 4$ are labelled "susceptible".
- The 69 isolates with MIC $\geq 8$ are labelled "resistant".

Predicted distribution of susceptible (S) and resistant (R) isolates from the model, with actual data shown in grey:

```{r}
ggplot(AC.CA_2.1.MM.mle$pred.n[[2]] %>% 
	 	mutate(MIC=factor(High), Pred.S = round(Pred.1,1), Pred.R = round(Pred.2,1), .keep="none") %>%
	 	pivot_longer(2:3,names_to="Category", values_to="N")) + 
	geom_col(aes(x=MIC,y=N, fill=Category)) +
	geom_vline(xintercept=7.5)+
	theme(axis.text.x = element_text(angle=75,hjust=1)) +
	geom_col(data=AC.CA_2.1.MM.mle$pred.n[[2]] %>% transmute(MIC=factor(High), N=N), aes(x=MIC, y=N), fill=NA, col=grey(0.5))
	
```


## Amoxicillin

```{r}
ggplot(mic.data.long %>% filter(Antibiotic=="Amoxicillin"), aes(x=as.factor(MIC))) +
	geom_bar(aes(fill = Genotype_EMH)) +
	theme(axis.text.x = element_text(angle=75,hjust=1))
```


```{r, include=F}
Amoxicillin.range <- read_excel("S epidermidis MIC data.xlsx",3)$`Amoxicillin` %>% sort()
Amoxicillin.MM.mle <- fit.multi.mix(mic.data$Amoxicillin, Amoxicillin.range, K=3)
```

```{r}
Amoxicillin.MM.mle$cut.offs
```

Proposed ECOFF for Amoxicillin: 5.66.

- The 31 isolates with MIC $\leq 4$ are labelled "susceptible".
- The 107 isolates with MIC $\geq 8$ are labelled "resistant".

Predicted distribution of susceptible (S) and resistant (R) isolates from the model, with actual data shown in grey:

```{r}
ggplot(Amoxicillin.MM.mle$pred.n[[3]] %>% 
	 	mutate(MIC=factor(High), Pred.S = round(Pred.1+Pred.2,1), Pred.R = round(Pred.3,1), .keep="none") %>%
	 	pivot_longer(2:3,names_to="Category", values_to="N")) + 
	geom_col(aes(x=MIC,y=N, fill=Category)) +
	geom_vline(xintercept=5.5)+
	theme(axis.text.x = element_text(angle=75,hjust=1)) +
	geom_col(data=Amoxicillin.MM.mle$pred.n[[3]] %>% transmute(MIC=factor(High), N=N), aes(x=MIC, y=N), fill=NA, col=grey(0.5))
	
```


## Penicillin

```{r}
ggplot(mic.data.long %>% filter(Antibiotic=="Penicillin"), aes(x=as.factor(MIC))) +
	geom_bar(aes(fill = Genotype_EMH)) +
	theme(axis.text.x = element_text(angle=75,hjust=1))
```


```{r, include=F}
Penicillin.range <- read_excel("S epidermidis MIC data.xlsx",3)$`Penicillin` %>% sort()
Penicillin.MM.mle <- fit.multi.mix(mic.data$Penicillin, Penicillin.range, K=3)
```

```{r}
Penicillin.MM.mle$cut.offs
```

Proposed ECOFF for Penicillin: 11.31.

- The 52 isolates with MIC $\leq 8$ are labelled "susceptible".
- The 86 isolates with MIC $\geq 16$ are labelled "resistant".

Predicted distribution of susceptible (S) and resistant (R) isolates from the model, with actual data shown in grey:

```{r}
ggplot(Penicillin.MM.mle$pred.n[[3]] %>% 
	 	mutate(MIC=factor(High), Pred.S = round(Pred.1+Pred.2,1), Pred.R = round(Pred.3,1), .keep="none") %>%
	 	pivot_longer(2:3,names_to="Category", values_to="N")) + 
	geom_col(aes(x=MIC,y=N, fill=Category)) +
	geom_vline(xintercept=6.5)+
	theme(axis.text.x = element_text(angle=75,hjust=1)) +
	geom_col(data=Penicillin.MM.mle$pred.n[[3]] %>% transmute(MIC=factor(High), N=N), aes(x=MIC, y=N), fill=NA, col=grey(0.5))
	
```


## Cefoxitin

```{r}
ggplot(mic.data.long %>% filter(Antibiotic=="Cefoxitin"), aes(x=as.factor(MIC))) +
	geom_bar(aes(fill = Genotype_EMH)) +
	theme(axis.text.x = element_text(angle=75,hjust=1))
```


```{r, include=F}
Cefoxitin.range <- read_excel("S epidermidis MIC data.xlsx",3)$`Cefoxitin` %>% sort()
Cefoxitin.MM.mle <- fit.multi.mix(mic.data$Cefoxitin, Cefoxitin.range, K=3)
```

```{r}
Cefoxitin.MM.mle$cut.offs
```

Proposed ECOFF for Cefoxitin: 90.5.

- The 84 isolates with MIC $\leq 64$ are labelled "susceptible".
- The 54 isolates with MIC $\geq 128$ are labelled "resistant".

Predicted distribution of susceptible (S) and resistant (R) isolates from the model, with actual data shown in grey:

```{r}
ggplot(Cefoxitin.MM.mle$pred.n[[3]] %>% 
	 	mutate(MIC=factor(High), Pred.S = round(Pred.1+Pred.2,1), Pred.R = round(Pred.3,1), .keep="none") %>%
	 	pivot_longer(2:3,names_to="Category", values_to="N")) + 
	geom_col(aes(x=MIC,y=N, fill=Category)) +
	geom_vline(xintercept=7.5)+
	theme(axis.text.x = element_text(angle=75,hjust=1)) +
	geom_col(data=Cefoxitin.MM.mle$pred.n[[3]] %>% transmute(MIC=factor(High), N=N), aes(x=MIC, y=N), fill=NA, col=grey(0.5))
	
```


## Clavulanic Acid

```{r}
ggplot(mic.data.long %>% filter(Antibiotic=="Clavulanic_acid"), aes(x=as.factor(MIC))) +
	geom_bar(aes(fill = Genotype_EMH)) +
	theme(axis.text.x = element_text(angle=75,hjust=1))
```


```{r, include=F}
Clavulanic_acid.range <- read_excel("S epidermidis MIC data.xlsx",3)$`Clavulanic acid` %>% sort()
Clavulanic_acid.MM.mle <- fit.multi.mix(mic.data$Clavulanic_acid,Clavulanic_acid.range, K=3)
```

```{r}
Clavulanic_acid.MM.mle$cut.offs
```


Proposed ECOFF for Clavulanic Acid: 45.2

- The 32 isolates with MIC $\leq 32$ are labelled "susceptible".
- The 106 isolates with MIC $\geq 64$ are labelled "resistant".

Predicted distribution of susceptible (S) and resistant (R) isolates from the model, with actual data shown in grey:

```{r}
ggplot(Clavulanic_acid.MM.mle$pred.n[[2]] %>% 
	 	mutate(MIC=factor(High), Pred.S = round(Pred.1,1), Pred.R = round(Pred.2,1), .keep="none") %>%
	 	pivot_longer(2:3,names_to="Category", values_to="N")) + 
	geom_col(aes(x=MIC,y=N, fill=Category)) +
	geom_vline(xintercept=6.5)+
	theme(axis.text.x = element_text(angle=75,hjust=1)) +
	geom_col(data=Clavulanic_acid.MM.mle$pred.n[[3]] %>% transmute(MIC=factor(High), N=N), aes(x=MIC, y=N), fill=NA, col=grey(0.5))
	
```

